Summary of Severe Weather Events Statistics across the USA (1993-2011)
========================================================

**P. Paquay**

# Synopsis

This study uses a Storm database from the The U.S. National Oceanic and Atmospheric Administration (NOAA) to determine which types of weather events are most harmful and which types have the greatest economic consequences across the USA.

To realize this study we did a lot of data cleaning (dates in Date format, property and crop damages in a computable format, subsetting on dates from 1993, subsetting on US states, ...). We also grouped the weather event types in seven categories for ease of readability as the number of unique values is too large to manage without these categories.

The results are presented in three different plots and six different tables.

# Data processing

## Locales

First we need to set the locales to the English language.

```{r cache = TRUE, echo = TRUE}
Sys.setlocale(locale = "en_GB.UTF-8")
```

## Software environment

The details about the software environment used for this analysis is given by the sessionInfo() function.

```{r cache = TRUE, echo = TRUE}
sessionInfo()
```

## Reading data

This phase consists simply in downloading the .csv.bz2 file with the download.file() function and reading the .csv file into a data frame named "storm.data".

```{r cache = TRUE, echo = TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2", "StormData.csv.bz2", method = "curl")
storm.data <- read.csv(bzfile("StormData.csv.bz2"), stringsAsFactors = FALSE)
```

The "storm.data" data frame has `r nrow(storm.data)` rows and `r ncol(storm.data)` columns.

## Cleaning data

We begin by creating a new variable "date" in the "storm.data" dataset by coercing the "BGN_DATE" variable into a Date format variable.

```{r cache = TRUE, echo = TRUE}
storm.data$date <- as.Date(storm.data$BGN_DATE, format = "%m/%d/%Y")
```

Then we create a new variable "prop.damage" in the "storm.data" dataset by using the "PROPDMG" as the coefficient and "PROPDMGEXP" as the exponent of the base 10 power (with B = 9, M = 6, K = 3, H = 2). We proceed in a similar way to create the variable "crop.damage".

```{r cache = TRUE, echo = TRUE}
storm.data$prop.damage <- rep(0, nrow(storm.data))
storm.data$prop.damage <- ifelse(storm.data$PROPDMGEXP == "B", storm.data$PROPDMG * 1e+09, ifelse(storm.data$PROPDMGEXP %in% c("M", "m"), storm.data$PROPDMG * 1e+06, ifelse(storm.data$PROPDMGEXP == "K", storm.data$PROPDMG * 1000, ifelse(storm.data$PROPDMGEXP %in% c("H", "h"), storm.data$PROPDMG * 100, ifelse(storm.data$PROPDMGEXP %in% seq("0":"9"), storm.data$PROPDMG * 10^as.numeric(as.character(storm.data$PROPDMGEXP)), storm.data$PROPDMG)))))

storm.data$crop.damage <- rep(0, nrow(storm.data))
storm.data$crop.damage <- ifelse(storm.data$CROPDMGEXP == "B", storm.data$CROPDMG * 1e+09, ifelse(storm.data$CROPDMGEXP %in% c("M", "m"), storm.data$CROPDMG * 1e+06, ifelse(storm.data$CROPDMGEXP == "K", storm.data$CROPDMG * 1000, ifelse(storm.data$CROPDMGEXP %in% c("H", "h"), storm.data$CROPDMG * 100, ifelse(storm.data$CROPDMGEXP %in% seq("0":"9"), storm.data$CROPDMG * 10^as.numeric(as.character(storm.data$CROPDMGEXP)), storm.data$CROPDMG)))))
```

## Creating categories

We will use seven categories as suggested by the "Summary of Natural Hazard Statistics for 2009 in the United States" from the National Weather Service (NWS) :
- *Convection* (lightning, tornado, thunderstorm and hail).
- *Extreme Temperatures* (cold and heat).
- *Flood* (flash flood and river flood).
- *Marine* (coastal storm, tsunami and rip current).
- *Tropical Cyclones* (tropical storm and hurricane).
- *Winter* (winter storm, ice and avalanche).
- *Other* (drought, dust storm, dust devil, ...).

To create these categories we will use several regular expressions namely "ind.convection", "ind.ext.temp", "ind.flood", "ind.marine", "ind.trop.cycl" and "ind.winter".

```{r cache = TRUE, echo = TRUE}
ind.convection <- "\\bL\\S+?G\\b|(NADO)|(\\bTOR\\S+?O\\b|(\\bFUN))|THUNDERSTORM|TSTM|(WIND)|(WND)|HAIL"
ind.ext.temp <- "COLD|HEAT|HYPERTHERMIA|HYPOTHERMIA|LOW TEMPERATURE|RECORD HIGH|RECORD LOW|Record temperature|RECORD WARM|Temperature record|UNSEASONABLY COOL|UNSEASONABLY HOT|UNUSUAL WARMTH|UNUSUAL/RECORD WARMTH|UNUSUALLY WARM|VERY WARM|WARM WEATHER|WARM DRY CONDITIONS|RECORD COLD"
ind.flood <- "(\\bFL\\S+?D)|RAIN|PRECIP|SHOWER"
ind.marine <- "^COASTAL(\\s)?STORM$|TSUNAMI|^RIP CUR"
ind.trop.cycl <- "HURRICANE|TROPICAL STORM"
ind.winter <- "(SNOW)|(ICE)|(ICY)|(FREEZ)|(WINT)|AVALAN|FROST|LOW TEMP|BLIZZARD"
```

Below you'll find the values of the "EVTYPE" variable placed into the *Convection* category.

```{r cache = TRUE, echo = TRUE}
sort(unique(grep(ind.convection, storm.data$EVTYPE, ignore.case = TRUE, value  = TRUE)))
```

Below you'll find the values of the "EVTYPE" variable placed into the *Extreme Temperature* category.

```{r cache = TRUE, echo = TRUE}
sort(unique(grep(ind.ext.temp, storm.data$EVTYPE, ignore.case = TRUE, value  = TRUE)))
```

Below you'll find the values of the "EVTYPE" variable placed into the *Flood* category.

```{r cache = TRUE, echo = TRUE}
sort(unique(grep(ind.flood, storm.data$EVTYPE, ignore.case = TRUE, value  = TRUE)))
```

Below you'll find the values of the "EVTYPE" variable placed into the *Marine* category.

```{r cache = TRUE, echo = TRUE}
sort(unique(grep(ind.marine, storm.data$EVTYPE, ignore.case = TRUE, value  = TRUE)))
```

Below you'll find the values of the "EVTYPE" variable placed into the *Tropical Cyclones* category.

```{r cache = TRUE, echo = TRUE}
sort(unique(grep(ind.trop.cycl, storm.data$EVTYPE, ignore.case = TRUE, value  = TRUE)))
```

Below you'll find the values of the "EVTYPE" variable placed into the *Winter* category.

```{r cache = TRUE, echo = TRUE}
sort(unique(grep(ind.winter, storm.data$EVTYPE, ignore.case = TRUE, value  = TRUE)))
```

## Grouping events into categories

As some events are listed in more than one category (e.g. "SNOW/COLD" which may be in the *Extreme Temperatures* or *Winter* categories), we need to fix some some precedence in the created categories. This hierarchy is suggested by the NWS and is ordered as :

1. *Convection*.
2. *Extreme Temperatures*.
3. *Flood*.
4. *Marine*.
5. *Tropical Cyclones*.
6. *Winter*.
7. *Other*.

So we now create a new factor variable named "category" as follows.

```{r cache = TRUE, echo = TRUE}
storm.data$category <- rep(0, nrow(storm.data))
storm.data$category <- ifelse(grepl(ind.convection, storm.data$EVTYPE, ignore.case = TRUE), 1, ifelse(grepl(ind.ext.temp, storm.data$EVTYPE, ignore.case = TRUE), 2, ifelse(grepl(ind.flood, storm.data$EVTYPE, ignore.case = TRUE), 3, ifelse(grepl(ind.marine, storm.data$EVTYPE, ignore.case = TRUE), 4, ifelse(grepl(ind.trop.cycl, storm.data$EVTYPE, ignore.case = TRUE), 5, ifelse(grepl(ind.winter, storm.data$EVTYPE, ignore.case = TRUE), 6, 7))))))
storm.data$category <- factor(storm.data$category)
levels(storm.data$category) <- c("convection", "extreme.temp", "flood", "marine", "tropical.cycl", "winter", "other")
```

## Restricting date range and locations

The date range of the "storm.data" dataset is given below.

```{r cache = TRUE, echo = TRUE}
c(min(storm.data$date), max(storm.data$date))
```

We see that the weather events go as far back as 1950. However, when we examine the date range for each category we find out that only the *Convection* events go back as far as 1950, the other categories reach only as far back as 1993.

```{r cache = TRUE, echo = TRUE}
min(subset(storm.data, category == "convection")$date)
min(subset(storm.data, category != "convection")$date)
```

So, to have coherent date ranges across categories, we will limit the date ranges to 1993 to 2011.

```{r cache = TRUE, echo = TRUE}
storm.data.restricted <- subset(storm.data, date >= "1993-01-01")
```

We will also limit our dataset to the fifty states of the USA as given in the "state.abb" vector.

```{r cache = TRUE, echo = TRUE}
storm.data.restricted <- subset(storm.data.restricted, STATE %in% state.abb)
```

# Results

## Fatalities

Below you'll find a stacked bar plot of the fatalities for each state and each category of events. To achieve this we need a new data frame "sum.fatalities" that computes the number of fatalities for each state and each category of events with the help of the agggregate() function.

```{r cache = TRUE, echo = TRUE, fig.width = 12}
library(ggplot2)
sum.fatalities <- aggregate(storm.data.restricted$FATALITIES, list(category = storm.data.restricted$category, state = storm.data.restricted$STATE), sum)
colnames(sum.fatalities)[3] <- "fatalities"
ggplot(sum.fatalities, aes(x = state, y = fatalities, fill = category)) + geom_bar(stat = "identity") + ggtitle("Fatalities by state and by category") + labs(x = "State", y = "Fatalities") + theme(plot.title = element_text(size = 20), legend.position = "bottom") + scale_fill_discrete("Weather Event", labels = c("Convection", "Extreme Temp", "Flood", "Marine", "Tropical cyclones", "Winter", "Other"))
```

We're able to see very clearly that the state of Illinois is the more severely affected by casualties (this is mainly due to the *Extreme Temperatures* type of events).

If we take a look at the fatalities in the entire USA, we may note that it is the *Convection* type of events that is responsible of the biggest part of fatalities due to weather events.

```{r cache = TRUE, echo = TRUE}
sum.fatalities.USA <- aggregate(sum.fatalities$fatalities, list(category = sum.fatalities$category), sum)
colnames(sum.fatalities.USA)[2] <- "fatalities"
sum.fatalities.USA[order(sum.fatalities.USA$fatalities, decreasing = TRUE), ]
```

To conclude this subsection, we give a table listing the category of events responsible of the greatest number of fatalities for each state.

```{r cache = TRUE, echo = TRUE}
library(plyr)
rank.fatalities <- ddply(sum.fatalities, "state", subset, fatalities == max(fatalities))
rank.fatalities <- rank.fatalities[c("state", "category", "fatalities")]
rank.fatalities
```

## Injuries

As in the previous subsection you'll find below a stacked bar plot of the injuries for each state and each category of events. To achieve this we need again a new data frame "sum.injuries" that computes the number of injuries for each state and each category of events with the help of the agggregate() function.

```{r cache = TRUE, echo = TRUE, fig.width = 12}
sum.injuries <- aggregate(storm.data.restricted$INJURIES, list(category = storm.data.restricted$category, state = storm.data.restricted$STATE), sum)
colnames(sum.injuries)[3] <- "injuries"
ggplot(sum.injuries, aes(x = state, y = injuries, fill = category)) + geom_bar(stat = "identity") + ggtitle("Injuries by state and by category") + labs(x = "State", y = "Injuries") + theme(plot.title = element_text(size = 20), legend.position = "bottom") + scale_fill_discrete("Weather Event", labels = c("Convection", "Extreme Temp", "Flood", "Marine", "Tropical cyclones", "Winter", "Other"))
```

In that case, we're able to see very clearly that the state of Texas is the more severely affected by injuries (this is mainly due to the *Flood* type of events).

If we take a look at the injuries in the entire USA, we may note that it is again the *Convection* type of events that is responsible of the biggest part of injuries due to weather events.

```{r cache = TRUE, echo = TRUE}
sum.injuries.USA <- aggregate(sum.injuries$injuries, list(category = sum.injuries$category), sum)
colnames(sum.injuries.USA)[2] <- "injuries"
sum.injuries.USA[order(sum.injuries.USA$injuries, decreasing = TRUE), ]
```

To conclude this subsection, we give a table listing the category of events responsible of the greatest number of injuries for each state.

```{r cache = TRUE, echo = TRUE}
rank.injuries <- ddply(sum.injuries, "state", subset, injuries == max(injuries))
rank.injuries <- rank.injuries[c("state", "category", "injuries")]
rank.injuries
```


## Material damages

As in the previous subsections you'll find below a stacked bar plot of the material damages for each state and each category of events. To achieve this we need again a new data frame "sum.damage" that computes the amount of property damages and crop damages combined for each state and each category of events with the help of the agggregate() function.

```{r cache = TRUE, echo = TRUE, fig.width = 12}
library(scales)
storm.data.restricted$material.damage <- (storm.data.restricted$crop.damage + storm.data.restricted$prop.damage) / 1000000
sum.damage <- aggregate(storm.data.restricted$material.damage, list(category = storm.data.restricted$category, state = storm.data.restricted$STATE), sum)
colnames(sum.damage)[3] <- "damages"
ggplot(sum.damage, aes(x = state, y = damages, fill = category)) + geom_bar(stat = "identity") + ggtitle("Damages (in million $) by state and by category") + labs(x = "State", y = "Damages (in million $)") + theme(plot.title = element_text(size = 20), legend.position = "bottom") + scale_fill_discrete("Weather Event", labels = c("Convection", "Extreme Temp", "Flood", "Marine", "Tropical cyclones", "Winter", "Other")) + scale_y_continuous(labels = comma)
```

In that case, we're able to see very clearly that the state of California is the more severely hit by economic consequences (this is mainly due to the *Flood* type of events).

If we take a look at the material damages in the entire USA, we may note that it is again the *Flood* type of events that have the greatest economic consequences due to weather events.

```{r cache = TRUE, echo = TRUE}
sum.damage.USA <- aggregate(sum.damage$damages, list(category = sum.damage$category), sum)
colnames(sum.damage.USA)[2] <- "damages"
sum.damage.USA[order(sum.damage.USA$damages, decreasing = TRUE), ]
```

To conclude this subsection, we give a table listing the category of events that have the greatest economic consequences for each state.

```{r cache = TRUE, echo = TRUE}
rank.damage <- ddply(sum.damage, "state", subset, damages == max(damages))
rank.damage <- rank.damage[c("state", "category", "damages")]
rank.damage
```
